From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-06 Pi+/paper/reversible1.tex
Revise section 3

>>>>

Using categorical language, the denotational semantics for $\PiLang$ is using the category of finite sets and functions
$\SetFin$. However, we only use the bijective functions for the semantics, which means, we use the groupoid $\BFin =
\term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and finite
products $(\unit, \times)$, which we use to interpret our types. In $\BFin$, the coproducts and products restrict to
additive and multiplicative symmetric monoidal structures, respectively, making $\BFin$ a symmetric rig groupoid.

### 2021-07-06 Pi+/paper/ufin.tex
Revise section 3

The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are finite sets
and whose morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
1-combinators as isomorphisms in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their
interpretations are extensionally equal in~$\BFin$. In other words, this groupoid is strict: two equal permutations are
extensionally identified, without an explicit witness for the equality. In order to establish completeness, we want to
quote back from the semantics to the syntax. Specifically, given a morphism in $\BFin$, we want to produce a
1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator
witnessing the equality of the quoted 1-combinators.

Our first step is therefore to expose the implicit equalities by \emph{weakening} the groupoid $\BFin$, i.e., by
replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
associativity equational conditions are replaced by higher invertible cells for the left and right units of composition
and for associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}. However, although weakening the groupoid
exposes some implicit equalities, it does not, by itself, give us a handle to characterize such equalities. One elegant
way to acquire such an ability is to ensure that the groupoid is ``univalent,'' as this provides a way to characterize
the semantic equivalences using the elements of the identity type, i.e., using objects whose equality is equipped with
an induction principle. We will therefore work within a HoTT metatheory which provides an ambient univalent universe of
types and introduce the novel idea of a \emph{univalent subuniverse} to carve out $\UFin$: a weakened, univalent,
version of the groupoid $\BFin$.

@@

%% In this section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.
%% Using the identity type for morphisms, a 1-groupoid in HoTT

IS THAT FOOTNOTE NECESSARY ??? We should avoid using footnotes as much as possible.\footnote{This is a locally-strict
  $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that is, truth-value
  enriched.}

@@

We review some basic concepts and notation from Homotopy Type Theory that we use in the appendix.~\footnote{We use the
language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional
Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for propositional
truncation and set quotients. All arguments will hold in a Cubical Type Theory as well, such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.}
For more details about HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

To recap, types in HoTT have a weak groupoid structure given by the identity type, where points are given by terms of
the type, and higher morphisms are given by the iterated identity type. Functions between types are functors between
groupoids, and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to
\UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the
structure of a fibration (see the appendix for details).

@@

The $\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure
of $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.

@@

The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent fibrations.

@@

\begin{definition}[Universe]
  A universe \`{a} la Tarski is given by the following pieces of data,
  \begin{itemize}
    \item a code $U : \UU$,
    \item a decoding type family $\El : U \to \UU$.
  \end{itemize}
  If $\El$ is univalent, $U$ is a univalent universe.
\end{definition}

\begin{definition}[Subuniverse]
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
  for every $X$. A subuniverse generated by $P$ is given by its total space $U \defeq \dsum{X:\UU}{P(X)}$ and $\El
  \defeq \fst$.
\end{definition}

\begin{proposition}[Univalent Subuniverse]
  Subuniverses generated by predicates are univalent.
\end{proposition}

@@

The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is subtle. We
define a type to be (Bishop) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.
\todo{Cite Brent Yorgey's thesis?}

\begin{definition}[$\Fin$]
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
    \end{aligned}
    \qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
\end{definition}

\begin{definition}[$\isFin$]
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
  \[
    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
  Note that the natural number $n$ need not be truncated, as justified below.
\end{definition}

@@

\begin{definition}
  The univalent subuniverse of \emph{all finite types} is given by
  \[
    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
  \]
  We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$.
\end{definition}

We also consider a different way of thinking about $\UFin$, using the subuniverse $\BAut[T]$.

\begin{definition}[$\BAut$]
  The predicate $P(X) \defeq \Trunc[-1]{X \id T}$ picks out exactly those types that are merely equal to $T$, and this
  generates the subuniverse
  \[
    \BAut[T] \defeq \Sub{T}.
  \]
  We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$
\end{definition}

Using $\BAut$, we can talk about types that have a specified cardinality, for example, the subuniverse of 2-element sets
is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give the denotational semantics for a 1-bit reversible
language~\cite{caretteReversibleProgramsUnivalent2018}.

\begin{definition}
  For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is, $\UFin \eqv
    \displaystyle\sqcup_{n:\Nat} \UFin[n]$.
\end{definition}

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$ is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
the automorphism group of $T$.

@@

\begin{corollary}
  $\UFin$ is a pointed, connected, 1-groupoid. For every $n:\Nat$, \( \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]] \).
\end{corollary}

The loopspace of a pointed type automatically has the structure of a group, that is, with $\refl_{\pt}$ for the neutral
element, path composition for the group multiplication, and path inverse for the group inverse operation. The group
axioms are then given by the higher paths corresponding to groupoid laws. We have shown that loops in $\UFin$ exactly
encode the automorphism group $\Aut[\Fin[n]]$ for every $n$. This is a general technique called delooping, where a group
can be identified with a 1-object groupoid, internally in HoTT.~\footnote{This technique also allows defining higher
groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}

@@

The groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative one, and the
multiplicative tensor product distributes over the additive one. To construct these, we first define some equivalences
on $\Fin$, and some general type isomorphisms.

@@

We lift these equivalences to $\UFin$ giving it the additive and multiplicative symmetric monoidal structures, $(F_0,
  \sqcup)$ and $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and
the braiding isomorphism $\mathcal{B}_{X,Y}$. These isomorphisms satisfy the Mac Lane coherence conditions for symmetric
monoidal categories, that is, the triangle, pentagon, and hexagon identities, and the symmetry of the braiding, upto
2-paths in $\UFin$.

>>>>

% First - recap the previous section
% Second - say what was wrong
To recap, the denotational semantics of $\PiLang$ interprets types as objects in $\BFin$, 1-combinators as isomorphisms,
and we prove that for every pair of 1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal.
The groupoid $\BFin$ is strict, since the collection of isomorphisms is a set, that is, a discrete category. There is no
explicit witness for the equality of two isomorphisms, since we can decide by evaluating two bijections whether they're
equal. To be able to establish completeness for $\PiLang$, we want a witness for this equality, so that we can quote
back to the syntax and produce a 2-combinator witnessing the equality of the corresponding 1-combinators.

% The denotational semantics for $\PiLang$ in the previous section uses the symmetric rig groupoid structure of $\BFin$,
% whose objects are finite sets, whose morphisms as bijections between finite sets, and whose symmetric monoidal
% structures are induced by the operations $\sqcup$ and $\times$. The

% two equal permutations are extensionally identified, without an explicit witness for the equality. In order to
% establish completeness, we want to quote back from the semantics to the syntax. Specifically, given a morphism in
% $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we
% want to produce a 2-combinator witnessing the equality of the quoted 1-combinators.

% Third - say how are we fixing it
First, we observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
homotopies -- we \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells for
them. We work in Homotopy Type Theory (Univalent Foundations) -- it provides a proof-relevant, constructive metatheory
to get a handle on these equalities, and, it provides a rich internal language for describing weak groupoids, using the
"types are weak $\infty$-groupoids" correspondence.

% To get a handle on these equalities, we will work in a proof-relevant metatheory, that is, HoTT, and then we will
% be able to write functions out of them back into the syntax.

% HoTT provides a rich internal language for describing weak groupoids.

% The idea is that we move from (discrete) sets to proof-relevant $\hSet$s, where the type of equalities between two
% points in a set is a proposition, that is, a prop-enriched groupoid.

% Our first step is therefore to expose the implicit equalities by \emph{weakening} the groupoid $\BFin$, i.e., by
% replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
% associativity equational conditions are replaced by higher cells for the left and right units of composition and for
% associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
% horizontal composition, or whiskering as detailed in appendix~\ref{app:twocat}.

% Fourth - expain the tool we use for fixing it
Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
gives the (higher) morphisms. The groupoid we're interested in has types as points, type equivalences for 1-cells, and
higher homotopies for higher cells. This is the groupoid structure for the universe type $\UU$, since the identity type
on types can be characterised as type equivalences (by \emph{univalence}). But, we only want to carve out a
\emph{subuniverse} of \emph{finite types}, still satisfying univalence, to get the groupoid structure. In this section,
we formally define \emph{univalent subuniverses}, and proceed to construct the particular instance for finite types,
$\UFin$ (\cref{def:ufin}).

% The way to describe such groupoid in HoTT is to construct a type which has the types themselves as points, i.e. a
% universe. Then, equivalences between those types can be characterized using elements of the equality type. For this to
% work, the universe in question has be a \emph{univalent subuniverse}.

% In HoTT, every type carries a groupoid structure, with terms for points, and the (iterated) identity type for (higher)
% morphisms. We're interested in the groupoid structure of the universe, which has types for points, equivalences for
% identities between them (by univalence), and higher homotopies for higher identitifications. We want a similar
% universe, whose only points are the finite types, that is, a subuniverse, and, equalities between them should
% correspond to equivalences of finite types -- the subuniverse should be univalent. To construct a univalent
% subuniverse starting from a univalent universe, we will introduce the concept of univalent fibrations.

% However, although weakening the groupoid exposes some implicit equalities, it does not, by itself, give us a handle to
% characterize such equalities. One elegant way to acquire such an ability is to ensure that the groupoid is
% ``univalent,'' as this provides a way to characterize the semantic equivalences using the elements of the identity
% type, i.e., using objects whose equality is equipped with an induction principle. We will therefore work within a HoTT
% metatheory which provides an ambient univalent universe of types and introduce the novel idea of a \emph{univalent
% subuniverse} to carve out $\UFin$: a weakened, univalent, version of the groupoid $\BFin$ which is then extended with
% the appropriate commutative rig structure. The main technical result of this section that is exploited in the next
% section is that, for each finite type $T$, we have transferred the structure of the space of equivalences $T \eqv T$
% to the structure of paths in $\UFin$. Since the space equivalences forms the automorphism group of $T$, we can now use
% the tools of computational group theory together with the induction principle of paths and term rewriting theory to
% describe the paths in $\UFin$.

@@

%% In this section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT. % Using
%the identity type for morphisms, a 1-groupoid in HoTT

% IS THAT FOOTNOTE NECESSARY ??? We should avoid using footnotes as much as possible.\footnote{This is a locally-strict
%   $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that is, truth-value
%   enriched.}

@@

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type Theory as well, such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
In the appendix, we review some basic concepts and notation from the HoTT book that we use. For more details, we refer
the reader to the book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

% To recap, types in HoTT have a weak groupoid structure given by the identity type, where points are given by terms of
% the type, and higher morphisms are given by the iterated identity type. Functions between types are functors between
% groupoids, and type families, or functions to the universe, are indexed families of groupoids.

@@

Functions between types are functors between groupoids, and type families, or functions to the universe, are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$
which has a lifting operation giving it the structure of a fibration (see the appendix for details). The
$\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure of
$A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.

@@

The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent
fibrations~\cite*{kapulkinUnivalenceSimplicialSets2018,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.

@@

\begin{definition}[Universe]
  A universe \`{a} la Tarski is given by the following pieces of data,
  \begin{itemize}
    \item a code $U : \UU$,
    \item a decoding type family $\El : U \to \UU$.
  \end{itemize}
  If $\El$ is univalent, we call $(U,\El)$ a \emph{univalent} universe.
\end{definition}

\begin{proposition}[Univalent Subuniverse]
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
  for every $X$. The fibration ${\fst : \dsum{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a univalent
  subuniverse
  \[
    \UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)
  \]
\end{proposition}

@@

The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is
subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness -- a type is finite if it is
merely equivalent to a finite set (\cref{def:finite-set}), and we define the predicate $\isFin$ (\cref{def:isfin}) to do
so.

\begin{definition}[$\Fin$]
  \label{def:finite-set}
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.
\end{definition}

\begin{definition}[$\isFin$]
  \label{def:isfin}
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
  \[
    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
  Note that the natural number $n$ need not be truncated, as justified below.
\end{definition}

@@

Since $\isFin[X]$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$.

\begin{definition}
  \label{def:ufin}
  The univalent subuniverse of \emph{all finite types} is given by
  \[
    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
  \]
  We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$.
\end{definition}

While $\UFin$ has \emph{all} the finite types, we're also interested in constructing a subuniverse of finite types of a
specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.

\begin{definition}[$\BAut$]
  The predicate $P(X) \defeq \Trunc[-1]{X \id T}$ picks out exactly those types that are merely equal to $T$, and this
  generates the subuniverse
  \[
    \BAut[T] \defeq \Sub{T}.
  \]
  We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$.
\end{definition}

Using $\BAut$, we can talk about types that are equivalent to a finite set of specified cardinality, for example, the
subuniverse of 2-element sets is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give the denotational semantics for a 1-bit reversible
programming language~\cite{caretteReversibleProgramsUnivalent2018}.

\begin{definition}
  For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is, $\UFin \eqv
    \dsum{n:\Nat}{\UFin[n]}$.
\end{definition}

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$ is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
the group of automorphisms of $T$.

@@

\begin{corollary}
  $\UFin[n]$ is a pointed, connected, 1-groupoid for every $n:\Nat$, and \( \loopspace[\UFin[n],F_{n}] \eqv
  \Aut[\Fin[n]] \). $\UFin$ is a 1-groupoid with connected components for every $n:\Nat$.
\end{corollary}

The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
path composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are
then given by the higher paths corresponding to groupoid laws. We have shown that loops in $\UFin$ exactly encode the
automorphism group $\Aut[\Fin[n]]$ for every $n$. This is a general technique called \emph{delooping}, where a group can
be identified with a 1-object groupoid, internally in HoTT.~\footnote{This technique also allows defining higher groups
  in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}

@@

Similar to $\BFin$, the groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative one,
and the multiplicative tensor product distributes over the additive one. To construct these, we first state and prove
some equivalences on $\Fin$, and some general type isomorphisms. Then we simply lift these equivalences to $\UFin$, by
the univalence principle.

@@

\begin{proposition}
  $\UFin$ has a symmetric rig structure, with the additive and multiplicative symmetric monoidal structures given by
  $(F_0, \sqcup)$ and $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$,
  $\alpha_{X,Y,Z}$, and the braiding isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy
  the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle, pentagon, and hexagon
  identities, and the symmetry of the braiding, upto 2-paths in $\UFin$.
\end{proposition}
